The Eclipse Integrated Computational Environment

### 2019-07-23 src/content.tex
Addressed comments from Charlie.
Signed-off-by: Jay Jay Billings <billingsjj@ornl.gov>

\begin{itemize}
\item
  modeling and simulation activities can be described in a succinct
  workflow model (see ``Workflow Model'');
\item
  an architecture for such a workflow system can satisfy the model of
  workflows in an extensible way (see ``Software Architecture''); and 
\item
  such a system is applicable to a suite of problems in energy science,
  including virtual battery simulations and additive manufacturing, 
  among others (see ``Illustrative Examples'').
\end{itemize}

@@

\textbf{Creating input} is the process of describing the physical model
or state of a system that will be simulated. This could include creating
an input file(s) or making calls to an external process to configure a
running program. In most situations, a computational scientist will
modify existing input or create new input from a template. ``Input''
generally includes run time parameters for the simulation framework
(e.g., tolerances); configuration options (e.g., data locations, output
locations, module configurations); properties of the materials to be
simulated; and a discretization of the simulation space (e.g., mesh,
grid, particle distribution). The collection of all required input can
be quite large and may go by many names, including ``input set,''
``input package,'' ``problem,'' or, simply, ``input.'' Often, the set of
input files will be described in a ``main'' input file that acts as a
kind of manifest to describe---and provide links to---all necessary
information for a given problem.

@@

Local jobs in ICE are executed using standard Java system calls. Remote jobs are launched only through SSH connections on remote machines. This includes direct SSH command execution on clusters and proxy connections through a pilot service on large Leadership-class supercomputers. Services such as Globus GRAM and Bosco are not supported, but the SSH command execution includes extensive support for numerous batch and queuing systems. The Eclipse Parallel Tools Platform (PTP) is used to create all remote SSH connections, regardless of the target machine \cite{tibbitts_integrated_2009}.

In this work, it should be assumed---unless otherwise noted---that
``executing a job'' includes monitoring that job in one or more ways,
possibly including real-time updates to visualizations. It is also important to note that executing a job is not the same as executing a workflow. Executing jobs specifically refers to launching simulations, whereas executing a workflow could be something different such as generating input or post-processing results. This is an important distinction because ICE's workflows are all executed locally, but simulations and work may be distributed remotely depending on the implementation of the workflow plugin.

\textbf{Analyzing results} includes executing special jobs to transform
data in one or more prescribed ways and producing artifacts with
scientific significance from the transformed data. This could include, for
example, post-processing results and visualizing the new data with
dedicated visualization tools. For many types of scientific computing,
this includes viewing the results of a simulation on a mesh or grid and
extracting publication-quality images or movies from that data. Other
cases might include analyzing results in preparation for follow-on
simulations or performing feature extraction, classification, or
activities for machine learning and data mining.

@@

\textbf{Modifying code} is not typically considered a part of a
scientific computing workflow. However, modeling and simulation use
cases often require users to explicitly modify code before execution or to issue special re-build instructions. This is true, for example, with the computational fluid dynamics code Nek5000, which requires modification of code before compilation and special build instructions using the \textit{makenek} shell script
\cite{the_nek5000_team_nek5000_2014}.
Many scientists consider ``their workflow'' to be re-running
software after modifications for purely exploratory purposes. This might
be required if the model that the author is modifying cannot be
configured directly as part of the input but can be easily accomplished by manipulating the source code.

@@

ICE's model of workflows differs significantly from many other
efforts in workflow science because it defines workflows in terms of high-level activities that are meant to guide workflow implementers and users. Many other workflow models in the literature define a workflow as
a collection of computing processes. For example, Yu and Buyya define
grid workflows as ``a collection of tasks that are processed on
distributed resources in a well-defined order to accomplish a specific
goal'' \cite{yu_taxonomy_2005}. Others, such as Pizzi et al., subscribe to
similar definitions \cite{pizzi_aiida:_2016}. This ``process'' view is
acceptable where the workflow is static and does not require additional
human input or ``human in the loop'' behavior after all the initial
human input is provided. However, workflows within ICE are fully interactive with regular callbacks to humans. It is simpler to discuss ``activities'' than it is to create a distinction between ``human processes'' and ``computer processes.''
Focusing on ``activities'' over processes (human or computer) also has
the benefit of removing concrete elements such as hardware or software properties that distract from details of workflows and workflow 
management systems. That is, considerations such as memory usage and 
raw performance are important, but questions about the abstract workflow 
or what the workflow management system should do are far more important in this context.

There are other workflow engines that have their own concepts of activities, with various similarities and differences compared to ICE's model. For example, Taverna has a concept of activities \cite{wolstencroft_taverna_2013}. Activities in Taverna are defined as invocable activities within a workflow, which is closely related to a lower-level class in ICE called Actions (see \ref{item}). On the other hand, ICE's activities loosely describe what can be done conceptually and the workflows that are implemented provide the Actions that carry out these activities.

ICE and Taverna can both natively execute Java code, but ICE 2.0 cannot natively invoke Web Service Definition Language (WSDL) web services. ICE uses regular Java-based libraries to interact with web services, such as Jersey, and requires that developers write the code to do this directly. Shim services in Taverna are provided in ICE as separate services in the framework, but are otherwise very similar.

Chiron, an algebraic workflow engine, has a concept of activities that is closer to ICE's than Taverna's, but still very strongly related to implementation \cite{ogasawara_chiron_2014}. Activities in Chiron are combinations of the program or expression to execute, plus the input schema, plus the output schema. This is a close match to the design of the Item class in ICE. ICE's Items (see Section \ref{item}) implement the high-level activities, such as launching a job or creating a model. The major difference between Chiron's activities and ICE's Items appears to be scope. A Chiron activity such as Map or Reduce would normally be implemented as an Action in ICE, whereas an Item in ICE may be tasked with executing multiple Actions to enact a workflow.

One significant difference between ICE and many other systems is that it uses a language-based approach to defining workflows, but with a standard language, Java, as opposed to a custom workflow definition language. Pizzi et al. do this with Aiida and Python, and the Fireworks Workflow Engine also uses Python. The advantage of this approach is that compilation is not a problem. ICE also differs from other workflow engines because it is specifically focused on modeling and simulation instead of grid workflows. Grid workflows are almost always defined as workflows based on trees without humans in the loop, but modeling and simulation workflows are much more interactive and undirected \cite{billings_toward_2017}.

@@

ICE was specifically created to address hands-on workflows for scientists, as in Section \ref{illustrative-examples}. Users download and
execute ICE locally, and ICE in turn orchestrates local or remote workflows as
required. It provides a comprehensive workbench for modeling and
simulation that includes tools for workflows, visualization, data
management, and software development.

@@

\begin{figure}[htbp]
\centering
\baseIncludegraphics{pubs/ice-softwarex-2017/src/images/ICE-arch-1.jpg}
\caption{High-level architecture of ICE showing the relationships and rough division of responsibility between the five primary components of ICE. The ICE Core is responsible for the workflow management and delegations instructions and data to the Item and Data Structures, respectively. User input is received from the client or through headless interactions, while persistence of large amounts of data is delegated to a persistence service.}
\label{highlevel-arch}
\end{figure}

Workflows and tasks in ICE are not explicitly treated as trees, or
directed acyclic graphs (DAGS), as is common with grid workflow tools
\cite{yu_taxonomy_2005}. Instead, ICE's design is inspired by representational
state transfer (REST) and the workflow engine is implemented as services in an Eclipse Rich Client Platform (RCP) application
\cite{fielding_architectural_2000} \cite{mcaffer_eclipse_2010}.

Figure \ref{highlevel-arch} shows the five primary components of ICE and their relationships to each other. Users initiate requests to create, edit, update, or delete workflows from the ``ICE Client'' (the workbench) or through headless interactions using web or language programming interfaces. The list of available workflows
that can be created is provided dynamically to the ICE Client by
the ``ICE Core,'' which acts as a server and is the primary component responsible for workflow orchestration. Workflow information is
provided dynamically because it often changes at run time based on 
both the configuration of available workflow components in the registry and on
persisted workflows that users have saved in their workspace. Workflow task enactment and execution is performed by services called ``Items.'' Information about workflows is provided to the Client by the Core
through common data structures rooted in a stateless ``Form'' tree. Forms describe the workflow and provide
all the necessary information to understand \emph{what} should process
the workflow (but not \emph{how} it should be processed). Once users
modify the description of the workflow in the Form to provide their
specific details, the Client dispatches a request to the Core to modify
and/or process the workflow. The Core then uses the information from the Form
to perform a service lookup to identify the correct Item to process the workflow.

@@

\begin{figure}[htbp]
\centering
\baseIncludegraphics{pubs/ice-softwarex-2017/src/images/ICE-arch-2.jpg}
\caption{The architecture of the ICE Core and its dependencies. The ICE Core class handles communications between workflow Items and common data structures, but delegates construction of workflow Items to builders.}
\label{core-arch}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\baseIncludegraphics{pubs/ice-softwarex-2017/src/images/ICE-arch-3.jpg}
\caption{The class hierarchy of ICE Items showing the relationships between the Item, its subclasses, and services that it uses. Vibe Launcher and MOOSE Launcher are examples of special Job Launch Items for executing jobs, and their applications are discussed in section \ref{illustrative-examples}.}
\label{item-arch}
\end{figure}

Each workflow Item in ICE is an independent service, and each
workflow type is a subclass of the Item base class or another Item subclass. Items are registered dynamically
through a service registry in ICE (see Section \ref{framework}) and
provide collections of workflow data, Forms, for the ICE Core and Client. Since ICE's design is
highly object oriented, it is easiest to think of the Item class as a
description of an abstract workflow and an Item object (an instance of
the class) as a concrete workflow with all required execution details
specified in its Form.

@@

\begin{table*}[t]
\begin{tabularx}{\textwidth}{|l|X|l|}
\hline
Class & Class Description & Object Description\tabularnewline\hline
Item & Java class with code to execute an abstract workflow. Provides a
Form. & Concrete workflow executor.\tabularnewline\hline
Form & Description and template of the data needed for the Item to
process the workflow. & User-modified workflow data.\tabularnewline\hline
Action & Java class for executing a specific task in the workflow. Used
by Items. & Concrete workflow task executor.\tabularnewline\hline
\end{tabularx}
\caption{Class Descriptions for Items, Forms, and Actions}
\end{table*}

@@

This design is very important. First, it means that all workflows in
ICE, regardless of their actual details and functions, can only behave in
a specific set of known and predictable ways, and this predictability 
simplifies the way the Core manages and interacts with Items. Second, by
formalizing state and error checks, ICE explicitly delineates which workflows 
can be executed from those that must receive additional configuration. 
Finally, it allows developers implementing Items and
Actions to specify by contract what is required before proceeding to the
next task, processing the workflow, or declaring a successful execution.

@@

ICE handles persistence using a ``persistence provider.'' The
default persistence provider uses JAX-RS to write Forms to 
XML~\cite{burke_restful_2010}. In principle, other persistence providers 
could replace the XML-based provider since it is registered as a dynamic 
service and a JAX-P based provider has been used in the past.

@@

ICE is an Eclipse Rich Client Platform (RCP) application
\cite{mcaffer_eclipse_2010} and has a plugin architecture based on 
Equinox, which is the reference implementation of the Open Service Gateway Initiative (OSGi)
framework specification \cite{mcaffer_osgi_2010}. It is most appropriate to think of Eclipse ICE as an alternative flavor of any other Eclipse development environment because like those environments it is just a collection of Eclipse plugins with a well-defined purpose and brand. ICE uses more than 1,200 additional packages from the Eclipse ecosystem to provide services like language
support and visualization. Each unique element of ICE described
in this work---including the Client, Core, Items, and data structures---is 
provided as plugins to Equinox. Most plugins are managed
dynamically and provided as services that can be obtained as dynamic OSGi Declarative Services. File I/O in ICE, with only a few exceptions,
interacts with the RCP's Resources plugin and the standard Eclipse Workspace. This also includes remote resources that are managed with Eclipse PTP.

ICE plugins are created used normal Eclipse development tools and installed into either the running instance of ICE or published in a newly built version. In either case, the running platform of ICE can only use the plugins that are actually installed and at least a small amount of code must be written in Java for each plugin. This extension method may seem odd to the novice user or developer, especially if they do not know Java very well, but
this design was chosen because it is an easy, fast, and precise way to create workflows.

ICE provides built-in development tools to automatically generate empty but ready to compile classes (``stubs'') of plugins that can then be installed into the framework as services. These tools are based on other code generation tools in the Eclipse platform. This
``self-hosting'' makes it possible for new users to create complex,
sophisticated workflows very quickly because they are not required to
know how to interact with the framework. Development of new workflows is streamlined because ICE provides a
rich API that exposes all workflow management functionality, documentation, tutorials, and tools to further simplify the process. These are shared through the self-hosted environment by code completion tools, help menus, and documentation overlays. The close proximity of the workflow environment with the development platform for modeling and simulation codes is also beneficial because, as a self-hosted entity, workflows can be co-developed with the code.

@@

There are many situations where configuring workflows graphically or serially is
unacceptable, such as when a very large number of workflows will be
executed or the type of information required is very fine grained. In
these cases, it is often necessary to provide scripts to the workflow
management system. ICE includes the Eclipse Advanced Scripting
Environment (EASE)~\cite{pontesegger_eclipse_2015} for scripting because it
provides a way to script Eclipse RCP projects natively in Javascript,
Jython, and Python. This also makes it possible to extend the
environment by adding Items in these languages. Specifically, scripts in Javascript and Python are written and executed in simple shell that is part of the workbench. Calling other languages, such as C or Fortran, is typically accomplished by adding workflow Items for the executables in these languages and performing local workflow job launches. C languages routines have been called from ICE through the Java Native Interface (JNI), but this is not a technique that is widely used with the platform because of its complexity.

@@

Solid-state sintering of parts printed using binder jetting
significantly increases part strength by decreasing the part porosity
and eliminating voids. However, this process does cause the resulting
product to shrink and warp from its original layout. An ideal near-net-shape 
process would combine binder jetting with solid-state sintering and account
for warpage and deformation in the part's design phase. 
Figure \ref{binder} shows an ICE-based workbench for performing simulations of
this process with visualizations of the pre- and post-simulation
properties of a central body with eight cantilevers. The primary
deformation in this type of geometry is bending or drooping of the
cantilevers due to sintering and thermal creep.

@@

The impact of software tools like ICE is difficult to quantify. However,
there are several examples of when ICE has significantly assisted its
development team and others.

One pressing area of interest and impact is that of interoperability
between workflow systems. There have been significant prior efforts in combining
large workflow management systems (e.g., Mandal et al.
\cite{mandal_integrating_2007}), but the end goal of gaining the greatest
advantage by using the best capabilities from multiple systems remains unrealized.
ICE's unique perspective on workflows and its well-defined API make it
possible to integrate multiple systems in a straightforward way. This
allows it to connect to other workflow environments, such as Triquetrum,
quite easily \cite{brooks_introducing_2016}. Triquetrum, like Kepler in
\cite{mandal_integrating_2007}, is a Ptolemy-based workflow engine
\cite{brooks_triquetrum:_2015}.

It is widely known that tools that enable researchers to be more productive tend
to improve the pursuit of new research. The high extensibility of
ICE and the tools that it combines from the larger Eclipse ecosystem
have made it possible for researchers on the development team to quickly
deploy new simulation environments for their research problems. Other tools created
with ICE might not invent something radically new, but they tend to
streamline interactions with those tools. Many ICE users, and certainly
the development team, have experienced improvements in their software
development efforts because of the tools that ICE provides or have learned
new technologies because access to new tooling was as simple as
installing more plugins through the Eclipse Marketplace.

@@

A new cloud-based development tool based on ICE is under development by
RNET Technologies, Inc., out of Dayton, Ohio, in response to a Small Business Innovation 
Research award from the US Department of Energy. This web-based version of 
ICE will continue ICE's support for nuclear energy and will integrate with cloud 
computing solutions like Amazon Web Services and ORNL's Compute and Data
Environment for Science (CADES). Additionally, while ICE has not directly led to
any ``spin-off'' companies, ICE source code has been used
in two spin-off projects: (1) EAVP (mentioned earlier) for advanced
visualizations and (2) the Eclipse January project for scientific data
structures \cite{graham_eclipse_2016}. ICE was also one of the founding
projects of the Eclipse Science Working Group.

@@

The primary resource for information on ICE is the project 
website~\cite{billings_eclipse_2016}. The ``Resources'' menu includes links to
detailed tutorials and user documentation. Examples of how to create
new workflow Items are available at
\url{https://github.com/eclipse/ice/tree/master/org.eclipse.ice.demo}.
Examples of how to use the scripting engine are available at
\url{https://github.com/eclipse/ice/tree/master/examples}. ICE also includes an
extensive suite of unit, integration, and UI tests, which are
also excellent examples of how to work with the platform. Tutorial and demonstration videos are available on YouTube at \url{https://goo.gl/nxCzRD}.

@@

Modeling and simulation workflows for scientists differ
greatly from those of experimentalists or those who primarily interact with
grid-based workflow management systems. The Eclipse ICE effort described here 
has been used to address interdisciplinary problems in 
modeling and simulation for energy science. Of particular interest are the
differences in architecture between a workflow management system focused
on modeling and simulation compared with those focused on grids. ICE's broad
applicability across many topics in energy science suggests that
there are opportunities for these systems in general. Finally, one 
interesting avenue of future exploration is coupling or integrating ICE 
with other workflow tools such as Aiida, Triquetrum, Kepler, and Pegasus,
which would make it possible to combine the best of grid and modeling workflows and
simulation workflows to address greater challenges.

\section*{Acknowledgments}\label{acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

The authors are grateful for the assistance and support of the following
people and institutions without whom this work would not have been
possible. This includes many people who directly contributed to the
project, either in its early days as ``NiCE'' or once it moved to
Eclipse, including Ronald Allen, Andrew Belt, David E.
Bernholdt, Tim Bohn, Erica Grant, Mike Guidry, John M. Hetrick III, Forest Hull, 
Sebastien Jourdain, JiSoo Kim, Allison Koenecke, Fangzhou Lin, Eric J. Lingerfelt, Greg
Lyon, Tony McCrary, Elizabeth Piersall, Neeti
Pokhriyal, Adrian Sanchez, Claire Saunders, Nick Stanish, Matthew Wang, and
Scott Wittenberg. The support of both the Nuclear Energy Advanced Modeling and Simulation (NEAMS) and Consortium for Simulation of Light Water Reactors (CASL) programs is also
greatly appreciated. The authors would like to acknowledge the special
contribution of the Eclipse Foundation, the Eclipse Community, the
Eclipse Science Working Group, and our many colleagues who use and
contribute to open-source projects in the Eclipse ecosystem.

Finally, the development team is especially grateful to Barney Maccabe,
David Pointer, and John Turner of ORNL for their
endless support and advocacy for this work.

>>>>

\begin{itemize}
\item
  modeling and simulation activities can be described in a succinct
  workflow model (see \S \ref{workflow-model}),
\item
  an architecture for such a workflow system can satisfy the model of
  workflows in an extensible way (see \S \ref{software-architecture}), and 
\item
  such a system is applicable to a suite of problems in energy science,
  including virtual battery simulations and additive manufacturing, 
  among others (see \S \ref{illustrative-examples}).
\end{itemize}

@@

\textbf{Creating input} is the process of describing the physical model
or state of a system that will be simulated. This could include creating
an input file(s) or making calls to an external process to configure a
running program. In most situations, a computational scientist will
modify existing input or create new input from a template. ``Input''
generally includes run time parameters for the simulation framework
(e.g., tolerances); configuration options (e.g., data locations, output
locations, module configurations); properties of the materials to be
simulated; and a discretization of the simulation space (e.g., mesh,
grid, particle distribution). The collection of all required input can
be quite large and can go by many names, including ``input set,''
``input package,'' ``problem,'' or, simply, ``input.'' Often, the set of
input files will be described in a ``main'' input file that acts as a
kind of manifest to describe---and provide links to---all necessary
information for a given problem.

@@

Local jobs in ICE are executed using standard Java system calls. Remote jobs
are launched only through SSH connections on remote machines. This includes
direct SSH command execution on clusters and proxy connections through a pilot
service on large leadership-class supercomputers. Services such as Globus GRAM
and Bosco are not supported, but the SSH command execution includes extensive
support for numerous batch and queuing systems. The Eclipse Parallel Tools
Platform (PTP) is used to create all remote SSH connections, regardless of the
target machine \cite{tibbitts_integrated_2009}.

In this work, it should be assumed---unless otherwise noted---that
``executing a job'' includes monitoring that job in one or more ways,
possibly including real-time updates to visualizations. It is also important to
note that executing a job is not the same as executing a workflow. Executing
jobs specifically refers to launching simulations, whereas executing a workflow
could be something different such as generating input or postprocessing
results. This is an important distinction because ICE's workflows are all
executed locally, but simulations and work can be distributed remotely
depending on the implementation of the workflow plugin.

\textbf{Analyzing results} includes executing special jobs to transform
data in one or more prescribed ways and producing artifacts with
scientific significance from the transformed data. This could include, for
example, postprocessing results and visualizing the new data with
dedicated visualization tools. For many types of scientific computing,
this includes viewing the results of a simulation on a mesh or grid and
extracting publication-quality images or movies from that data. Other
cases might include analyzing results in preparation for follow-on
simulations or performing feature extraction, classification, or
activities for machine learning and data mining.

@@

\textbf{Modifying code} is not typically considered a part of a
scientific computing workflow. However, modeling and simulation use
cases often require users to explicitly modify code before execution or to
issue special build instructions. This is true, for example, with the
computational fluid dynamics code Nek5000, which requires modification of code
before compilation and special build instructions using the \textit{makenek}
shell script \cite{the_nek5000_team_nek5000_2014}.
Many scientists consider ``their workflow'' to be re-running
software after modifications for purely exploratory purposes. This might
be required if the model the author is modifying cannot be
configured directly as part of the input but can be easily accomplished by
manipulating the source code.

@@

ICE's model of workflows differs significantly from many other
efforts in workflow science because it defines workflows in terms of high-level
activities that are meant to guide workflow implementers and users. Many other
workflow models in the literature define a workflow as a collection of
computing processes. For example, Yu and Buyya define grid workflows as ``a
collection of tasks that are processed on distributed resources in a
well-defined order to accomplish a specific goal'' \cite{yu_taxonomy_2005}.
Others, such as Pizzi et al., subscribe to similar definitions
\cite{pizzi_aiida:_2016}. This ``process'' view is acceptable where the
workflow is static and does not require additional human input or ``human in
the loop'' behavior after all the initial human input is provided. However,
workflows within ICE are fully interactive with regular callbacks to humans. It
is simpler to discuss ``activities'' than it is to create a distinction between
``human processes'' and ``computer processes.'' Focusing on activities over
processes (human or computer) also has the benefit of removing concrete
elements such as hardware or software properties that distract from details of
workflows and workflow management systems. That is, considerations such as
memory usage and raw performance are important, but questions about the
abstract workflow or what the workflow management system should do are far more
important in this context.

There are other workflow engines that have their own concepts of activities,
with various similarities and differences compared with ICE's model. For
example, Taverna has a concept of activities \cite{wolstencroft_taverna_2013}.
Activities in Taverna are defined as invocable activities within a workflow,
which is closely related to a lower-level class in ICE called Actions (see \S
\ref{item}). On the other hand, ICE's activities loosely describe what can be
done conceptually and the workflows that are implemented provide the Actions
that carry out these activities.

ICE and Taverna can both natively execute Java code, but ICE 2.0 cannot
natively invoke Web Service Definition Language (WSDL) web services. ICE uses
regular Java-based libraries to interact with web services, such as Jersey, and
requires that developers write the code to do this directly. Shim services in
Taverna are provided in ICE as separate services in the framework but are
otherwise very similar.

Chiron, an algebraic workflow engine, has a concept of activities that is
closer to ICE's than Taverna's but still very strongly related to
implementation \cite{ogasawara_chiron_2014}. Activities in Chiron are
combinations of the program or expression to execute, plus the input schema,
plus the output schema. This is a close match to the design of the Item class
in ICE. ICE's Items (see Section \ref{item}) implement the high-level
activities, such as launching a job or creating a model. The major difference
between Chiron's activities and ICE's Items appears to be scope. A Chiron
activity such as Map or Reduce would normally be implemented as an Action in
ICE, whereas an Item in ICE can be tasked with executing multiple Actions to
enact a workflow.

One significant difference between ICE and many other systems is that it uses a
language-based approach to defining workflows but with a standard language,
Java, as opposed to a custom workflow definition language. Pizzi et al. do this
with Aiida and Python, and the Fireworks Workflow Engine also uses Python. The
advantage of this approach is that compilation is not a problem. ICE also
differs from other workflow engines because it is specifically focused on
modeling and simulation instead of grid workflows. Grid workflows are almost
always defined as workflows based on trees without humans in the loop, but
modeling and simulation workflows are much more interactive and undirected
\cite{billings_toward_2017}.

@@

ICE was specifically created to address hands-on workflows for scientists, as
described in \S \ref{illustrative-examples}. Users download and execute ICE
locally, and ICE in turn orchestrates local or remote workflows as required. It
provides a comprehensive workbench for modeling and simulation that includes
tools for workflows, visualization, data management, and software development.

@@

\begin{figure}[htbp] \centering
\baseIncludegraphics{pubs/ice-softwarex-2017/src/images/ICE-arch-1.jpg}
\caption{High-level architecture of ICE showing the relationships and rough
division of responsibility between the five primary components of ICE. The ICE
Core is responsible for the workflow management and delegations instructions
and data to the Item and data structures, respectively. User input is received
from the client or through headless interactions, while persistence of large
amounts of data is delegated to a persistence service. The ``1'' and ``*''
represent the cardinality of the aggregation relationship.}
\label{highlevel-arch}
\end{figure}

Workflows and tasks in ICE are not explicitly treated as trees or
directed acyclic graphs (DAGS), as is common with grid workflow tools
\cite{yu_taxonomy_2005}. Instead, ICE's design is inspired by representational
state transfer (REST), and the workflow engine is implemented as services in an
Eclipse Rich Client Platform (RCP) application
\cite{fielding_architectural_2000} \cite{mcaffer_eclipse_2010}.

Figure \ref{highlevel-arch} shows the five primary components of ICE and their
relationships to each other. Users initiate requests to create, edit, update, or
delete workflows from the ``ICE Client'' (the workbench) or through headless
interactions using web or language programming interfaces. The list of available
workflows that can be created is provided dynamically to the ICE Client by the
``ICE Core,'' which acts as a server and is the primary component responsible
for workflow orchestration. Workflow information is provided dynamically because
it often changes at run time based on both the configuration of available
workflow components in the registry and on persisted workflows users have
saved in their workspace. Workflow task enactment and execution is performed by
services called ``Items.'' Information about workflows is provided to the Client
by the Core through common data structures rooted in a stateless ``Form'' tree.
Forms describe the workflow and provide all the necessary information to
understand \emph{what} should process the workflow (but not \emph{how} it should
be processed). Once users modify the description of the workflow in the Form to
provide their specific details, the Client dispatches a request to the Core to
modify and/or process the workflow. The Core then uses the information from the
Form to perform a service lookup to identify the correct Item to process the
workflow.

@@

\begin{figure}[htbp]
\centering
\baseIncludegraphics{pubs/ice-softwarex-2017/src/images/ICE-arch-2.jpg}
\caption{Architecture of the ICE Core and its dependencies. The ICE Core class
handles communications between workflow Items and common data structures but
delegates construction of workflow Items to builders.}
\label{core-arch}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\baseIncludegraphics{pubs/ice-softwarex-2017/src/images/ICE-arch-3.jpg}
\caption{The class hierarchy of ICE Items showing the relationships between the
Item, its subclasses, and services that it uses. Vibe Launcher and MOOSE
Launcher are examples of special Job Launch Items for executing jobs, and their
applications are discussed in \S \ref{illustrative-examples}.}
\label{item-arch}
\end{figure}

Each workflow Item in ICE is an independent service, and each workflow type is a
subclass of the Item base class or another Item subclass. Items are registered
dynamically through a service registry in ICE (see Section \ref{framework}) and
provide collections of workflow data ---Forms--- for the ICE Core and Client.
Since ICE's design is highly object oriented, it is easiest to think of the Item class
as a description of an abstract workflow and an Item object (an instance of the
class) as a concrete workflow with all required execution details specified in
its Form.

@@

\begin{table*}[t]
\begin{tabularx}{\textwidth}{|l|X|l|}
\hline
Class & Class Description & Object Description\tabularnewline\hline
Item & Java class with code to execute an abstract workflow. Provides a
Form. & Concrete workflow executor.\tabularnewline\hline
Form & Description and template of the data needed for the Item to
process the workflow. & User-modified workflow data.\tabularnewline\hline
Action & Java class for executing a specific task in the workflow. Used
by Items. & Concrete workflow task executor.\tabularnewline\hline
\end{tabularx}
\caption{Class descriptions for Items, Forms, and Actions.}
\end{table*}

@@

This design is very important. First, it means that all workflows in
ICE, regardless of their actual details and functions, can behave only in
a specific set of known and predictable ways, and this predictability 
simplifies the way the Core manages and interacts with Items. Second, by
formalizing state and error checks, ICE explicitly delineates which workflows 
can be executed from those that must receive additional configuration. 
Finally, it allows developers implementing Items and
Actions to specify by contract what is required before proceeding to the
next task, processing the workflow, or declaring a successful execution.

@@

ICE handles persistence using a ``persistence provider.'' The
default persistence provider uses JAX-RS to write Forms to 
XML~\cite{burke_restful_2010}. In principle, other persistence providers 
could replace the XML-based provider since it is registered as a dynamic 
service and a JAXP based provider has been used in the past.

@@

ICE is an Eclipse RCP application \cite{mcaffer_eclipse_2010} and has a plugin
architecture based on Equinox, which is the reference implementation of the Open
Service Gateway Initiative (OSGi) framework specification
\cite{mcaffer_osgi_2010}. It is most appropriate to think of Eclipse ICE as an
alternative flavor of any other Eclipse development environment because like
those environments it is just a collection of Eclipse plugins with a
well-defined purpose and brand. ICE uses more than 1,200 additional packages
from the Eclipse ecosystem to provide services like language support and
visualization. Each unique element of ICE described in this work---including the
Client, Core, Items, and data structures---is provided as plugins to Equinox.
Most plugins are managed dynamically and provided as services that can be
obtained as dynamic OSGi Declarative Services. File input/output in ICE, with
only a few exceptions, interacts with the RCP's Resources plugin and the
standard Eclipse Workspace. This also includes remote resources that are
managed with Eclipse PTP.

ICE plugins are created using normal Eclipse development tools and are installed
into either the running instance of ICE or published in a newly built version.
In either case, the running platform of ICE can use only the plugins that are
actually installed and at least a small amount of code must be written in Java
for each plugin. This extension method may seem odd to the novice user or
developer, especially if they do not know Java very well, but this design was
chosen because it is an easy, fast, and precise way to create workflows.

ICE provides built-in development tools to automatically generate empty but
ready to compile classes (``stubs'') of plugins that can then be installed into
the framework as services. These tools are based on other code generation tools
in the Eclipse platform. This ``self-hosting'' makes it possible for new users
to create complex, sophisticated workflows very quickly because they are not
required to know how to interact with the framework. Development of new
workflows is streamlined because ICE provides a rich API that exposes all
workflow management functionality, documentation, tutorials, and tools to
further simplify the process. These are shared through the self-hosted
environment by code completion tools, help menus, and documentation overlays.
The proximity of the workflow environment with the development platform
for modeling and simulation codes is also beneficial because, as a self-hosted
entity, workflows can be codeveloped with the code.

@@

There are many situations where configuring workflows graphically or serially is
unacceptable, such as when a very large number of workflows will be
executed or the type of information required is very fine grained. In
these cases, it is often necessary to provide scripts to the workflow
management system. ICE includes the Eclipse Advanced Scripting
Environment (EASE)~\cite{pontesegger_eclipse_2015} for scripting because it
provides a way to script Eclipse RCP projects natively in Javascript,
Jython, and Python. This also makes it possible to extend the
environment by adding Items in these languages. Specifically, scripts in
Javascript and Python are written and executed in simple shell that is part of
the workbench. Calling other languages, such as C or Fortran, is typically
accomplished by adding workflow Items for the executables in these languages
and performing local workflow job launches. C language routines have been
called from ICE through the Java Native Interface (JNI), but this is not a
technique that is widely used with the platform because of its complexity.

@@

Solid-state sintering of parts printed using binder jetting
significantly increases part strength by decreasing the part porosity
and eliminating voids. However, this process does cause the resulting
product to shrink and warp from its original layout. An ideal near-net-shape 
process would combine binder jetting with solid-state sintering and account
for warpage and deformation in the part's design phase. 
Figure \ref{binder} shows an ICE-based workbench for performing simulations of
this process with visualizations of the pre- and postsimulation
properties of a central body with eight cantilevers. The primary
deformation in this type of geometry is bending or drooping of the
cantilevers due to sintering and thermal creep.

@@

The impact of software tools like ICE is difficult to quantify. However, there
are several examples of ICE significantly assisting the development team and
others.

One pressing area of interest and impact is that of interoperability
between workflow systems. Significant prior efforts have been made to
combine large workflow management systems (e.g., Mandal et al.
\cite{mandal_integrating_2007}), but the end goal of gaining the greatest
advantage by using the best capabilities from multiple systems remains unrealized.
ICE's unique perspective on workflows and its well-defined API make it
possible to integrate multiple systems in a straightforward way. This
allows ICE to connect to other workflow environments, such as Triquetrum,
quite easily \cite{brooks_introducing_2016}. Triquetrum, like Kepler in
\cite{mandal_integrating_2007}, is a Ptolemy-based workflow engine
\cite{brooks_triquetrum:_2015}.

It is widely known that tools that enable researchers to be more productive tend
to improve the pursuit of new research. The high extensibility of
ICE and the tools that it combines from the larger Eclipse ecosystem
have made it possible for researchers on the development team to quickly
deploy new simulation environments for their research problems. Other tools created
with ICE might not invent something radically new, but they tend to
streamline interactions within those tools. Many ICE users, and certainly
the development team, have experienced improvements in their software
development efforts because of the tools that ICE provides or have learned
new technologies because access to new tooling was as simple as
installing more plugins through the Eclipse Marketplace.

@@

A new cloud-based development tool based on ICE is under development by
RNET Technologies, Inc., out of Dayton, Ohio, in response to a Small Business Innovation 
Research award from the DOE. This web-based version of 
ICE will continue ICE's support for nuclear energy and will integrate with cloud 
computing solutions like Amazon Web Services and ORNL's Compute and Data
Environment for Science (CADES). Additionally, although ICE has not directly led
to any ``spin-off'' companies, ICE source code has been used
in two spin-off projects: (1) EAVP (mentioned earlier) for advanced
visualizations and (2) the Eclipse January project for scientific data
structures \cite{graham_eclipse_2016}. ICE was also one of the founding
projects of the Eclipse Science Working Group.

@@

The primary resource for information on ICE is the project 
website~\cite{billings_eclipse_2016}. The ``Resources'' menu includes links to
detailed tutorials and user documentation. Examples of how to create
new workflow Items are available at
\url{https://github.com/eclipse/ice/tree/master/org.eclipse.ice.demo}.
Examples of how to use the scripting engine are available at
\url{https://github.com/eclipse/ice/tree/master/examples}. ICE also includes an
extensive suite of unit, integration, and user interface tests, which are
also excellent examples of how to work with the platform. Tutorial and
demonstration videos are available on YouTube at \url{https://goo.gl/nxCzRD}.

@@

Modeling and simulation workflows for scientists differ
greatly from those of experimentalists or those who primarily interact with
grid-based workflow management systems. The Eclipse ICE effort described here 
has been used to address interdisciplinary problems in 
modeling and simulation for energy science. Of particular interest are the
differences in architecture between a workflow management system focused
on modeling and simulation compared with systems focused on grids. ICE's broad
applicability across many topics in energy science suggests
there are opportunities for these systems in general. Finally, one 
interesting avenue of future exploration is coupling or integrating ICE 
with other workflow tools such as Aiida, Triquetrum, Kepler, and Pegasus,
which would make it possible to combine the best of grid and modeling workflows and
simulation workflows to address greater challenges.
